Improving Adaptive Seamless Designs through Bayesian optimization

### 2021-03-15 response_to_the_reviewers.md

# Response to the reviews for "Improving Adaptive Seamless Designs through Bayesian optimization"

We would like to thank all reviewers for their valuable and constructive comments on the paper.
We especially that Reviewer 1 for the thoughtful and in-depth feedback.

## Reviewer: 1

### Comments to the Author

> This is a well organised and written manuscript, describing a very thorough investigation of a reasonably interesting problem. Computational aspects of complex adaptive clinical trial designs have all too often been neglected. The speed gain proposed here is substantial, with no apparent loss in quality. The findings could help improve the uptake of adaptive seamless designs. I only have a few relatively minor suggestions that I hope the authors will find helpful.

### General

> 1) Power is used as a key performance metric, yet it lacks an explicit definition. The problem, of course, is that power is slightly more complicated in a multi-stage design with multiple arms (and therefore multiple hypotheses to be tested) than it is in a simple comparison of two groups. Given the scenario described in the motivating example, I would have assumed that power was defined as the probability of at least one (any) treatment group being demonstrated to be ‘better’ than control i.e. a ‘disjunctive’ or ‘minimal’ power definition (cf. Senn & Bretz 2007). However, on p. 10 it is stated that “the vector of treatment numbers for determining power counts rejections of 3 or 4 hypotheses”, implying a power definition more similar to ‘conjunctive’ or ‘maximal’ power.
>
> Reference: Senn S, Bretz F (2007) Power and sample size when multiple endpoints are considered. Pharmaceutical Statistics, 6(3), 161-170.

--> Tim Friede

>2) There is some inconsistent notation:
>
>a) It seems that n_treat is varyingly defined as the “total number of treatments” (p. 8, l. 29 and p. 9, l. 20), “total number of patients” (p. 8, l. 48 and p. 9, l. 14) and “allowed treatment number” (p. 12, l. 50 and p. 12, l. 51).

n_treat is the number of total treatments. 
If the patients are different in in both stages (i.e. each patient is only treated once), both numbers are the same.
However, we corrected the wording to avoid confusions.

> b) On p. 3 k is introduced as the running index for treatments (k = 1, …, K) but later on p. 9 k is the resolution of the search grid.

Thanks for catching that mistake. It is now fixed.

### Specific

>p. 2: For the benefit of readers less familiar with the term, the authors should explain at least briefly what they mean by “expensive black-box” optimisation problems.

We added a short explanation.

>p. 3 ff.: This may be implicitly clear but it should also be said explicitly that in the motivating example and throughout the remainder of the paper a normality assumption is made for all outcome data.

--> Tim? Gibt es diese assumption bei dir?
--> Mbo Teil habe ich übernommen.
We clarified the assumptions in chapter 4 and 5.

>p. 9: How were the values for the resolution of Grid and Grid Small (k=25 and k=7) selected?

We added an explanation and referenced previous studies, that used similar values.

> p. 10: Similarly, how was the number of randomly sampled points (16) for the initial design for BO chosen?

We added an explanation and a reference.

> p. 10: In section 5.4 ten additional stochastic simulations are mentions, which seems in conflict with the 20 additional stochastic simulations mentioned in other parts of the manuscript.

Thanks, for spotting that mistake. Indeed we use 20 stochastic repetitions.

> p. 10: It should be clarified whether the test level of 0.025 is one-sided.

--> Tim

> p. 11: Whilst I appreciate that the focus is on relative improvements in runtime, the authors should at least briefly describe the hardware setup with which these runtimes were achieved, to provide some context for the absolute runtimes.

We added the requested information.

> p. 12: It is stated that: “Only for the scenarios (effect: paper2, n_treat = 1000) and (effect: sigmoid, n_treat = 1000) Grid Small has a small advantage over BO and Grid.” I think it should be added that Grid Small also has an advantage over BO for (effect: paper, n_treat = 1000) and certainly for (effect: sigmoid, n_treat = 2000).

> p. 12: It says: “For example, for scenario effect: sigmoid, n_treat = 1000 Grid is superior to BO, but the difference is smaller than 0.25%.” This isn’t true for n_treat = 1000 where the difference between Grid and BO is around 2%, but rather for n_treat = 2000.

> p. 12: Apologies but I’m struggling to understand the following sentence: “As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2.” How can none of the treatments be used in stage 2 when on p. 8 it is specified that the number of treatments in stage 2 must always be between 2 and 5?

--> clarify

> p. 13: It is stated that: “Note, that the selection strategies thresh and all never selected powerful designs.” Why is that? Presumably, the former is true because the range of numerical values used for the threshold was chosen too narrow, whilst the latter shouldn’t be surprising at all, as it is the only non-adaptive design in the mix.

--> Maybe threshold is disadvantageous because it looks at the absolute values, which is not that adaptive even with optimization

> p. 14: What are the x-values referred to in the caption of Figure 4? No x is introduced in the notation, and Figure 4 itself plots y-values against r-values.

> p. 16: The citation of the technical report by Bischl et al. should include a URL if possible.

### Typos

> p. 9, l. 20: “in average”, which should read “on average”

Done.

> p. 9, l. 24: “in 8”, which should read “in (8)”

Done.

> p. 9, l. 53 and > p. 10, l. 14: “Mattern”, which should be spelt “Matérn”

> p. 10, l. 28: “Gird”

Done.

> p. 10, l. 39: “???”

Done.

> p. 12, l. 19: “Gris”

Done.

> p. 13, l. 48f.: remove closing bracket from “500)”

Done.

> p. 14, l. 55: “which conflicts the goal is to find the optimal” (grammar)

Done.

## Reviewer: 2

### Comments to the Author

> The authors propose to use Bayesian optimization (BO) to improve the efficiency of the design selection process in clinical trials. A set of parameters to be chose for the design optimization based on the power. The idea seem novel and could be a useful approach. However, there are several major issues and limitations of proposed methods: for example, power is often only part of measure of choosing design, other factors like study duration, and # of pts are also important. Also, the black-box function relay too much of the various parametric assumption of parameters, which need to be carefully decided to be more related to true clinical trial needs. Overall, I have the following comments for the authors.

--> In out method we keep the number of patients fixed
--> Tim Friede (assumption of parameters, true clinical trial needs)

### Major comments:

> In page 7 table 1, the effect size scenarios presented seem always assume 2nd stage effect size are higher than stage 1? The authors should also evaluate the vice visa situation, and consistent effect size case in simulation.

--> Argument: Falschrum
--> Often Realisitic scenario, also for the motivating example: First stage, second stage ...
--> Tim

> In page 8 table 2, the selection of “espilon” and “threshold” used in the arm selection should be based on the clinical meaningful cut, I am not sure how practical that they are part of parameters to be optimized? The authors should also evaluate with different setting.

--> Argument: Given a simulation setting we optimize the parameters eps and thr. This is independent of the application but of course the clinical application can give hints which simulation settings are adequate

> In section 5.2, the authors keep the total number of pts fixed, which seem not appropriate in practice, as overall power relying several factors, e.g. # of pts, the interim selection rule, stage 1 and stage 2 ratio. Often # of pts is also key goal of optimal design evaluation. I will strongly suggest # of pts should be a parameter in the black-box function, instead of fixing it to a constant value.

--> Argument: This will only lead to a result where more patients are selected. The idea is to run this optimization for different patient sizes, and then to decide which number of patients is required (also mention this in paper)

> In page 10 table 4, the number of simulation iteration is only 1000, which does not seem sufficient for complex optimization methods proposed by authors. I would suggest at least 5000 should be needed.

--> simulation iteration means that we evaluated 1000 times wheter the test detected a significant effect. this does not depend on the complexity of the optimization. As a matter of fact we use 100 optimization iterations which is sufficient as can be seen in plot ... where we reach convergence.

> Page 11 table 5, the authors present the average run time in hour, not sure how these information is helpful, as this depend on the what computing systems were used in the simulation. Please either add more information.

--> add computational information, relative values are still meaningfull

> Page 13, BO has issue to find optimal value when one parameters closer to the borders of search space. This is concerning and confirm the limitation of using too many assumptions for optimization, e.g. appropriate parameters range. I am not sure taking log-transform could solve this problem, could authors elaborate more details on this.  Additional simulation is necessary to confirm this.

--> log transform: good idea, outlook

> Given authors used COPD trial as motivating example, I would strongly suggest to add section of case study for the illustration of proposed methods.

--> in fact our effect set "paper" issues a case study where the effect sets are estimated on a true dataset. we do not propose a method to analyze data, but a method to select the best design given a known theoretical situation. (check that this is clear)

> Minor comments:

Page 8 line 29,  n_{treat} should be total # of pts, instead of “total number of treatment”.

### 2021-03-15 main.tex
Updates from Overleaf
fixes and literature

BO, often also called Model-based optimization, has been successfully applied to optimize expensive black-boxes in many scenarios.
Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
Various implementations of BO methods are widely used, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012}, and BoTorch~\citep{balandat_botorch_2020}.
For this work we use the implementation in the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017}.
This implementation has been successfully used to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
Also, BO is well suited for multi-criteria optimization.
Different adaptions of BO exist~\citep{horn_modelbased_2015} that return a set of non-dominated points.
Many of them are implemented in \texttt{mlrMBO} as well.
However, in this work we restrict ourselves to single-criterion optimization.

@@

\subsection{Acquisition function for noisy outcomes}
Since this work focuses on optimization of a stochastic simulation, we assume that our non-deterministic optimization problem can be formulated as follows:
\begin{equation}
\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta) + \varepsilon, \, \text{ with } \, \varepsilon \sim N(0, \sigma^2) %\text{ i.i.d.}
\end{equation}
In this setting, the expected improvement is not feasible, as shown in \citet{huang_global_2006}.
They propose to use the augmented expected improvement instead.
Therefore, in a first step, we calculate the \emph{effective best solution}:
\begin{equation}
  \theta^{\ast\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta) - c \cdot s(\theta),
\end{equation}
with $c$ as a tuning parameter that is usually set to 1 and $\mathcal{D}$ as the design that contains all previously evaluated values of $\theta$.
The effective best point is the pessimistic estimate of the best observed outcome so far.
In the final step, we calculate the augmented expected improvement~\citep{huang_global_2006} as follows:
\begin{multline}
  \label{eq:aei}
  \operatorname{AEI}(\theta) = \left( \hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast}) \right) \cdot  \Phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{s(\theta)} \biggr) \\
   + s(\theta) \cdot \phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{s(\theta)} \biggr) \cdot \biggl(1 - \underbrace{\frac{\sigma}{\sqrt{\sigma^2 + s^2(\theta)}}}_{\text{correction}}\biggr),
\end{multline}
where $\Phi$ and $\phi$ are the distribution and density function of the standard normal distribution, and $\sigma^2$ denotes the random error (nugget effect) in the Kriging model.
The formula is derived under the assumption, that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
This assumption is also reflected by using a Gaussian process regression as a surrogate.
The intuition in formula (\ref{eq:aei}) is that the first summand favors exploitation while the second summand favors exploration.
The correction term is necessary to avoid exploration in areas where the model uncertainty $s^2(\theta)$ equals the random error $\sigma^2$, because in this case further evaluations will not decrease the model uncertainty.

@@

Given the number of treatments $k_1$ and $k_2$ in the two stages and the total number of patients $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
\begin{align}
  \label{eq:stagec}
  %\begin{split}
  n_{\text{stage1}} & = 
  n_{\text{stage1}}\cdot \frac{n_{\text{treat}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} = 
  n_{\text{treat}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \nonumber \\
  & = n_{\text{treat}}\cdot
  \frac{r}{k_1\cdot
  r+k_2\cdot (1-r)} \ \text{and} \nonumber \\
  n_{\text{stage2}} & =
  n_{\text{treat}}\cdot
  \frac{1-r}{k_1\cdot
  r+k_2\cdot (1-r)}.
  %\end{split}
\end{align}
%For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
These formulas give the numbers of patients per treatment arm in the two stages, if $k_2$ is known. 
This is only the case for selection strategies 1, 2, and 3.

For the selection strategies \emph{eps} and \emph{thresh}, $k_2$ depends on the outcome of stage 1.
%choice of the \emph{thresh} or \emph{eps} parameter.
In these cases, for given values of $n_{\text{stage1}}$ and $\epsilon$ or $\tau$, respectively, we can obtain an estimate $\hat{k}_2$ of $k_2$ with a so-called calibration step.
For a set of values for $n_{\text{stage1}}$ in the interval $\left[ \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil, \lceil \frac{n_{\text{treat}}}{k_1} \rceil \right]$, we simulate stage 1 of the trial multiple times.
We calculate $n_{\text{stage2}} = (1-r)/r\cdot n_{\text{stage1}}$ from formula (\ref{eq:r}) and $\hat{k}_2$ as the average of the resulting values of $k_2$ for all simulations with a fixed value for $n_{\text{stage1}}$.
Then we select the value of $n_{\text{stage1}}$ such that the corresponding number of total patients is closest to $n_{\text{treat}}$.
More precisely, we minimize
\begin{align}
  \label{eq:targettreat}
  \begin{split}
  h(n_{\text{stage1}}) &:= \left( \Big(k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot %\underbrace{
  \frac{1-r}{r} \cdot n_{\text{stage1}}
  %}_{=n_{\text{stage2}}}
  \Big) - n_{\text{treat}} \right)^2
%  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
  \end{split}
\end{align}
with respect to $n_{\text{stage1}}$, where the estimate $\hat{k}_2$ is obtained as described above dependent on $n_{\text{stage1}}$. 
%$f$ is the simulator black-box from Equation~\ref{eq:bbox}, that gives us the estimated number of $\hat{k}_2$.
%This number is independent of the simulation of the second stage, so we can set $n_{\text{stage2}}$ to 1.
%The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.
The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.

The above procedure gives us the values for $n_{\text{stage1}}$ and $n_{\text{stage2}}$, allowing us in the following to find the optimal clinical trial design for a given \emph{effect set} and number of total treatments, by optimizing the parameter values for $r$, \emph{Selection strategy}, $\epsilon$ and $\tau$ through maximizing $\tilde f$ in~\ref{eq:bbox2}.

@@

\paragraph{Grid and Grid small}
We conduct an \emph{exhaustive grid search} with a resolution of $k$ points per real-valued dimension and one point for each categorical dimension.
Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of $k$.
All 6 selection strategies have the numeric parameter $r$.
The strategies \emph{epsilon} and \emph{thresh} both have an additional second parameter, $\epsilon$ or $\tau$, respectively.
Using $k=25$, we call this approach \textbf{Grid},
with in a total $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
Using a subset of this grid with $k=7$ points per dimension, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, we call the method \textbf{Grid Small}.
Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.

@@

\paragraph{BO (Model-based optimization) and BO Grid}
For applying BO, we first define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ (parameter configurations) is generated.
As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.

Kriging expects a numerical input, but our search space contains categorical parameters and is even hierarchical.
First, to transform the categorical parameters of the search space (see Table~\ref{tab:search_space}) into numerical values we use one-hot encoding.
Furthermore, our search space is hierarchical because the parameters $\epsilon$ and $\tau$ are only active for the respective selection strategies \emph{eps} and \emph{thresh}.
The hierarchical structure introduces inactive values for the configurations ($\theta$), i.e.\ if \emph{selection strategy} is set to \emph{eps} then the parameter $\tau$ of \emph{thresh} is inactive.
If a value is inactive, we set it to a value two times as high as the maximum in its active range (e.g.\ if $\tau$ is inactive it will be set to 20).
This ensures that within the original range the Kriging model is not or only minimally affected by the inactive values.
Note, that this trick only works because we did not scale the input variables to $[0,1]$ and because the estimated covariance of the Mattern$\frac{5}{2}$ kernel is approximately zero for distances greater than $4$.
With the above steps we ensure that the input for the Kriging is purely numeric.

As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:bayesian_optimization}.
We start BO with an initial design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.

@@

Additionally, we investigate how \emph{BO} performs when it is limited to return configurations that are within the grid introduced previously.
Therefore, we introduce \textbf{BO Grid}, which executes the optimization exactly as \emph{BO}, but the best configuration is mapped to the closest value on the grid. 
In other words, for \emph{BO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{BO}.
The comparison of the results of \emph{BO Gird} and \emph{Gird} will allow two possible conclusions:
First, if \emph{BO Grid} is worse than \emph{BO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
Second, if \emph{BO Grid} is better than \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
Summing up, there are two factors that can prevent \emph{Grid} from finding the optimal configuration: lack of resolution and noise of the outcome.

@@

The algorithms used in this study are implemented in R.
For the Bayesian optimization, the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017} is used.
For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $f$. 
The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.

% (TODO Not sure what ptest that means)  %jeweils Elementarhypothese 3, 4 verwerfen, gucken wir gar nicht drauf

\begin{table}[h]
  \caption{Parameter choices for the R function \texttt{treatsel.sim}}
  \label{tab:par_implement}
  \centering
  \begin{tabular}{lll}
  \hline
  Meaning of parameter & Parameter & Value \\
  \hline
  Number of simulation iterations & \texttt{nsim} & $1000$ \\
  Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
  Test level & \texttt{level} & $0.025$ \\
  Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
  \hline
  \end{tabular}
\end{table}

>>>>

BO, often also called Model-based optimization, has been successfully applied to optimize expensive black-boxes (i.e.\ functions that take long to evaluate and are not given in closed-form) in many scenarios.
Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
Various implementations of BO methods are widely used, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012}, and BoTorch~\citep{balandat_botorch_2020}.
For this work we use the implementation in the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017}.
This implementation has been successfully used to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
Also, BO is well suited for multi-criteria optimization.
Different adaptions of BO exist~\citep{horn_modelbased_2015} that return a set of non-dominated points.
Many of them are implemented in \texttt{mlrMBO} as well.
However, in this work we restrict ourselves to single-criterion optimization.

@@

The expected improvement, as well as other acquisition functions, are derived under the assumption that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
The standard Gaussian process regression, that is often used as surrogate, models the posterior as a Gaussian with the parameters \hat{\mu} and \sigma^2 under the assumption that $f$ is a realization of a Gaussian process.
In practice, this assumption often cannot be verified.
However, benchmarks show that the approach nevertheless works reliable for many practical problems~\citep{bischl_mlrmbo_2017,snoek_practical_2012}.

\subsection{Acquisition function for noisy outcomes}
Since this work focuses on optimization of a stochastic simulation, we assume that our non-deterministic optimization problem can be formulated as follows:
\begin{equation}
\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta) + \varepsilon, \, \text{ with } \, \varepsilon \sim N(0, \sigma^2) %\text{ i.i.d.}
\end{equation}
In this setting, the expected improvement is not feasible, as shown in \citet{huang_global_2006}.
They propose to use the augmented expected improvement instead.
Therefore, in a first step, we calculate the \emph{effective best solution}:
\begin{equation}
  \theta^{\ast\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta) - c \cdot s(\theta),
\end{equation}
with $c$ as a tuning parameter that is usually set to 1 and $\mathcal{D}$ as the design that contains all previously evaluated values of $\theta$.
The effective best point is the pessimistic estimate of the best observed outcome so far.
In the final step, we calculate the augmented expected improvement~\citep{huang_global_2006} as follows:
\begin{multline}
  \label{eq:aei}
  \operatorname{AEI}(\theta) = \left( \hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast}) \right) \cdot  \Phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{s(\theta)} \biggr) \\
   + s(\theta) \cdot \phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{s(\theta)} \biggr) \cdot \biggl(1 - \underbrace{\frac{\sigma}{\sqrt{\sigma^2 + s^2(\theta)}}}_{\text{correction}}\biggr),
\end{multline}
where $\Phi$ and $\phi$ are the distribution and density function of the standard normal distribution, and $\sigma^2$ denotes the random error (nugget effect) in the Kriging model.
The formula is derived under the assumption of a Gaussian posterior at each point $\theta$.
The intuition in formula (\ref{eq:aei}) is that the first summand favors exploitation while the second summand favors exploration.
The correction term is necessary to avoid exploration in areas where the model uncertainty $s^2(\theta)$ equals the random error $\sigma^2$, because in this case further evaluations will not decrease the model uncertainty.

@@

Given the number of treatments $k_1$ and $k_2$ in the two stages and the total number of treatments $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
\begin{align}
  \label{eq:stagec}
  %\begin{split}
  n_{\text{stage1}} & = 
  n_{\text{stage1}}\cdot \frac{n_{\text{treat}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} = 
  n_{\text{treat}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \nonumber \\
  & = n_{\text{treat}}\cdot
  \frac{r}{k_1\cdot
  r+k_2\cdot (1-r)} \ \text{and} \nonumber \\
  n_{\text{stage2}} & =
  n_{\text{treat}}\cdot
  \frac{1-r}{k_1\cdot
  r+k_2\cdot (1-r)}.
  %\end{split}
\end{align}
%For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
These formulas give the numbers of patients per treatment arm in the two stages, if $k_2$ is known. 
This is only the case for selection strategies 1, 2, and 3.

For the selection strategies \emph{eps} and \emph{thresh}, $k_2$ depends on the outcome of stage 1.
%choice of the \emph{thresh} or \emph{eps} parameter.
In these cases, for given values of $n_{\text{stage1}}$ and $\epsilon$ or $\tau$, respectively, we can obtain an estimate $\hat{k}_2$ of $k_2$ with a so-called calibration step.
For a set of values for $n_{\text{stage1}}$ in the interval $\left[ \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil, \lceil \frac{n_{\text{treat}}}{k_1} \rceil \right]$, we simulate stage 1 of the trial multiple times.
We calculate $n_{\text{stage2}} = (1-r)/r\cdot n_{\text{stage1}}$ from formula (\ref{eq:r}) and $\hat{k}_2$ as the average of the resulting values of $k_2$ for all simulations with a fixed value for $n_{\text{stage1}}$.
Then we select the value of $n_{\text{stage1}}$ such that the corresponding number of total treatments is closest to $n_{\text{treat}}$.
More precisely, we minimize
\begin{align}
  \label{eq:targettreat}
  \begin{split}
  h(n_{\text{stage1}}) &:= \left( \Big(k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot %\underbrace{
  \frac{1-r}{r} \cdot n_{\text{stage1}}
  %}_{=n_{\text{stage2}}}
  \Big) - n_{\text{treat}} \right)^2
%  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
  \end{split}
\end{align}
with respect to $n_{\text{stage1}}$, where the estimate $\hat{k}_2$ is obtained as described above dependent on $n_{\text{stage1}}$. 
%$f$ is the simulator black-box from Equation~\ref{eq:bbox}, that gives us the estimated number of $\hat{k}_2$.
%This number is independent of the simulation of the second stage, so we can set $n_{\text{stage2}}$ to 1.
%The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.
The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (on average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.

The above procedure gives us the values for $n_{\text{stage1}}$ and $n_{\text{stage2}}$, allowing us in the following to find the optimal clinical trial design for a given \emph{effect set} and number of total treatments, by optimizing the parameter values for $r$, \emph{Selection strategy}, $\epsilon$ and $\tau$ through maximizing $\tilde f$ in~(\ref{eq:bbox2}).

@@

\paragraph{Grid and Grid small}
We conduct an \emph{exhaustive grid search} with a resolution of $l$ points per real-valued dimension and one point for each categorical dimension.
Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of $l$.
All 6 selection strategies have the numeric parameter $r$.
The strategies \emph{epsilon} and \emph{thresh} both have an additional second parameter, $\epsilon$ or $\tau$, respectively.
For $l=25$, we call this approach \textbf{Grid},
with a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
We chose $l=25$ generously to obtain an estimate of the optimum that should be close to the true optimum.
Previous studies used half the resolution to assess the behavior of similar problems~\citep{friede_adaptive_2020}.
We define \textbf{Grid Small} to be a subset of this grid with $l=7$ points per dimension, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$.
This gives a number of evaluations that can be easily carried out in practice.

Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.

@@

\paragraph{BO (Model-based optimization) and BO Grid}
For applying BO, we first define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ (parameter configurations) is generated.
As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
We configure Kriging to apply the Matérn$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.
Note, that by doing so, we implicitly assume that $f$ is a realization of a Gaussian process and that each single outcome is a realization of a Gaussian.
However, as stated in Section~\ref{sec:bayesian_optimization}, we can expect the method to work well even if the assumption is slightly violated.

Kriging expects a numerical input, but our search space contains categorical parameters and is even hierarchical.
First, to transform the categorical parameters of the search space (see Table~\ref{tab:search_space}) into numerical values we use one-hot encoding.
Furthermore, our search space is hierarchical because the parameters $\epsilon$ and $\tau$ are only active for the respective selection strategies \emph{eps} and \emph{thresh}.
The hierarchical structure introduces inactive values for the configurations ($\theta$), i.e.\ if \emph{selection strategy} is set to \emph{eps} then the parameter $\tau$ of \emph{thresh} is inactive.
If a value is inactive, we set it to a value two times as high as the maximum in its active range (e.g.\ if $\tau$ is inactive it will be set to 20).
This ensures that within the original range the Kriging model is not or only minimally affected by the inactive values.
Note, that this trick only works because we did not scale the input variables to $[0,1]$ and because the estimated covariance of the Matérn$\frac{5}{2}$ kernel is approximately zero for distances greater than $4$.
With the above steps we ensure that the input for the Kriging is purely numeric.

As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:bayesian_optimization}.
We start BO with an initial design of $16$ randomly sampled points, following an established rule-of-thumb of $4$ points per dimension, which is also well in line with the recommendation of using around $10\%$ of your budget for the initial design~\cite{bossek_initial_2020}.
We allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.

@@

Additionally, we investigate how \emph{BO} performs when it is limited to return configurations that are within the grid introduced previously.
Therefore, we introduce \textbf{BO Grid}, which executes the optimization exactly as \emph{BO}, but the best configuration is mapped to the closest value on the grid. 
In other words, for \emph{BO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{BO}.
The comparison of the results of \emph{BO Grid} and \emph{Grid} will allow two possible conclusions:
First, if \emph{BO Grid} is worse than \emph{BO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
Second, if \emph{BO Grid} is better than \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
Summing up, there are two factors that can prevent \emph{Grid} from finding the optimal configuration: lack of resolution and noise of the outcome.

@@

The algorithms used in this study are implemented in R.
For the Bayesian optimization, the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017} is used.
For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses. %???
Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $f$. 
The 20 additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.

% (TODO Not sure what ptest that means)  %jeweils Elementarhypothese 3, 4 verwerfen, gucken wir gar nicht drauf
%FIXME: Ist mir immer noch nicht klar / nicht mehr klar?

\begin{table}[h]
  \caption{Parameter choices for the R function \texttt{treatsel.sim}}
  \label{tab:par_implement}
  \centering
  \begin{tabular}{lll}
  \hline
  Meaning of parameter & Parameter & Value \\
  \hline
  Number of simulation iterations & \texttt{nsim} & $1000$ \\
  Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
  Test level & \texttt{level} & $0.025$ \\ %FIXME: @Tim: One sided?
  Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
  \hline
  \end{tabular}
\end{table}
